Learning action models from {\em state-invariants}

### 2019-02-14 doc/ijcai-19-deductivelearning/deductivelearning.tex
updates added to ijcai19 paper on deductive learning

\begin{abstract}
This paper addresses the learning of \strips\ action models from {\em state-invariants} (i.e logic formulae that specify constraints about the possible states of a given domain). The benefit of exploiting {\em state-invariants} is two-fold, they constrain the space of possible action models and they can complete learning examples that are only partially observed. Our approach for learning \strips\ action from {\em state-invariants} is a {\em classical planning} compilation that is flexible to different kinds of input knowledge (e.g., partially observations of plan executions including partially observed intermediate states and/or actions). The experimental results demonstrate that, even at unfavorable scenarios where input observations are minimal (just an {\em initial state} and the {\em goals}), {\em state-invariant} are helpful to learn good quality \strips\ action models.
\end{abstract}

\section{Introduction}
\label{sec:introduction}

The specification of planning action models is a complex process that limits, too often, the application of {\em model-based planning} to real-world tasks~\cite{kambhampati:modellite:AAAI2007}. The {\em machine learning} of action models relieves this {\em knowledge acquisition bottleneck} of {\em model-based planning} and nowadays, there exists a wide range of effective approaches for learning action models~\cite{arora:amodels:ker2018}. Many of the most successful approaches for learning planning action models are however purely {\em inductive}~\cite{yang2007learning,pasula2007learning,mourao2010learning,zhuo2013action}, linking learning performance exclusively to the {\em amount} and {\em quality} of the input learning examples.

This paper addresses the learning of action models exploiting a different source of knowledge, {\em deductive} knowledge. Our approach leverages {\em state-invariants}, i.e. logic formulae that specify constraints about the possible states of a given domain, to cushion the negative impact of insufficient learning examples. Given an action model, state-of-the-art planners infer {\em state-invariants} from that model to reduce the search space and make the planning process more efficient~\cite{helmert2009concise}. In this paper we follow the opposite direction and leverage {\em state-invariants} to learn the planning action model. The benefit of learning action models from {\em state-invariants} is two-fold, {\em state-invariants} constrain the space of possible action models and can complete learning examples that are only partially observed.

Our approach for learning \strips\ action models from {\em state-invariants} is compile the learning task into a classical planning task. Our compilation is flexible to different kinds of input knowledge (e.g., partially/fully observations of actions of plan executions as well as partially/fully observed intermediate states) and outputs an action model that is {\em consistent} with the given input knowledge. The experimental results demonstrate that, even at unfavorable scenarios where input observations are minimal (just an {\em initial state} and the {\em goals}), {\em state-invariant} help to learn better \strips\ models.

\section{Background}
\label{sec:background}
This section formalizes the {\em classical planning model} we follow in this work and introduces the classical planning compilation for the learning of \strips\ action models~\cite{aineto2018learning}. Finally, the section formalizes {\em state-invariants}.

@@

\subsection{Learning action models with classical planning}
This work is built on top of the {\em classical planning compilation} for the learning \strips\ action models~\cite{aineto2018learning}. This compilation receives as input an empty model $\mathcal{M}$ (which only contains the action headers), and an observation of a plan execution $\mathcal{O}(\tau)$ (extensible to a set of observations). The compilation outputs a model $\mathcal{M'}$ that specifies the preconditions and effects of each action schema included in $\mathcal{M}$ such that the validation of $\mathcal{O}(\tau)$ following $\mathcal{M'}$ is successful; i.e., it holds $\rho(s_{i-1}^o,a_i)$ for every observed action of $\mathcal{O}(\tau)$ and $s_i^o=\theta(s_{i-1}^o,a_i)$ for every observed state of $\mathcal{O}(\tau)$.

Essentially, a solution plan to the classical planning problem that results from the compialtion is a sequence of: (a) \emph{insert actions} that insert preconditions and effects on the schemata of $\mathcal{M}$ to build $\mathcal{M'}$ and (b) \emph{apply actions} that validate the application of the $\mathcal{M'}$ model in the observation $\mathcal{O(\tau)}$.

To illustrate this, Figure~\ref{fig:plan-lplan} shows a solution to a classical planning problem resulting from the~\citeauthor{aineto2018learning}~\citeyear{aineto2018learning} compilation. In the initial state of that problem three  blocks ({\small\tt blockA}, {\small\tt blockB} and {\small\tt blockC}) are clear and on top of the table, the robot hand is empty. The problem goal is having the three-block tower {\tt blockA} on top of {\tt blockB} and {\tt blockB} on top of {\tt blockC}. The plan shows the {\em insert} actions for the {\tt\small stack} scheme (steps $00-01$ insert the preconditions, steps $05-10$ insert the effects), steps $02-04$ insert the preconditions of the {\tt\small pickup} scheme (while steps $10-13$ insert the effects of this scheme). Finally, steps $14-17$ is the plan postfix that applies the programmed action model to achieve the goals $G$ starting from $I$.

\begin{figure}[hbt!]
	{\tiny\tt
\begin{tabular}{ll}
		{\bf 00}:(insert\_pre\_stack\_holding\_v1) & {\bf 10}:(insert\_eff\_pickup\_clear\_v1) \\
		01:(insert\_pre\_stack\_clear\_v2) & 11:(insert\_eff\_pickup\_ontable\_v1)\\
                {\bf 02}:(insert\_pre\_pickup\_handempty) & 12:(insert\_eff\_pickup\_handempty)\\
                03:(insert\_pre\_pickup\_clear\_v1) & 13:(insert\_eff\_pickup\_holding\_v1)\\
                04:(insert\_pre\_pickup\_ontable\_v1) & {\bf 14}:(apply\_pickup blockB)\\
                {\bf 05}:(insert\_eff\_stack\_clear\_v1) & 15:(apply\_stack blockB blockC)\\
                06:(insert\_eff\_stack\_clear\_v2) & 16:(apply\_pickup blockA)\\
                07:(insert\_eff\_stack\_handempty) & 17:(apply\_stack blockA blockB) \\
                08:(insert\_eff\_stack\_holding\_v1) &  {\bf 18}:(validate\_1)\\
                09:(insert\_eff\_stack\_on\_v1\_v2) &             		 
\end{tabular}
}
	\caption{\small Plan computed when solving a problem output by the classical planning compilation.}
	\label{fig:plan-lplan}
\end{figure}

@@

{\em State invariants} is a kind of state-constraints useful for computing more compact state representations~\cite{helmert2009concise} or making {\em satisfiability planning} and {\em backward search} more efficient~\cite{rintanen2014madagascar,alcazar2015reminder}. Given a classical planning problem $P=\tup{F,A,I,G}$, a {\em state invariant} is a formula $\phi$ that holds at the initial state of a given classical planning problem, $I\models \phi$, and at every state $s$, built from $F$, that is reachable from $I$ by applying actions in $A$. For instance, Figure~\ref{fig:strongest-invariant} shows five clauses that define {\em state invariants} for the {\em blocksworld} planning domain~\cite{slaney2001blocks}.

@@

We define a {\em domain mutex} as a $\tup{p,q}$ predicates pair where both $p\in\Psi$ and $q\in\Psi$ are predicates that shape the set of fluents $F$ of a given planning problem and such that they satisfy the following formulae $p\rightarrow \neg q$ where the predicate variables are universally quantified. For instance, predicates $holding(x)$ and $clear(x)$ from the {\em blocksworld} are {\em domain mutex} while predicates $clear(x)$ and $ontable(x)$ are not ($\forall x\ clear(x)\leftrightarrow\neg ontable(x)$ does not hold for every possible {\em blocksworld} state). We pay special attention to {\em domain mutex} because they identify the {\em properties} of a given type of objects~\cite{fox:TIM:JAIR1998} and because they enable (1) effectively pruning of inconsistent \strips\ action models and (2) effective completion of partially observed states.

@@

\subsection{The sampling space}
We define a {\em learning example} $\mathcal{O}=\tup{s_0^o,s_1^o \ldots, s_m^o}$ as a sequence of partial states, except for the initial state $s_0^o$, which is fully observed. A partially observable state $s_i^o$ is one in which $|s_i^o| < |F|$; i.e., a state in which at least a fluent of $F$ is not observable. Note that this definition also comprises the case $|s_i^o| = 0$, when the state is fully unobservable. Each {\em observed} states comprises $[1,|F|]$ fluents. The observation can still miss intermediate states that are {\em unobserved} so transiting between two consecutive observed states in $\mathcal{O}$ may require the execution of more than a single action ($\theta(s_i^o,\tup{a_1,\ldots,a_k})=s_{i+1}^o$, where ${\small k\geq 1}$ is unknown but finite.

@@

\subsection{The learning task}
We define the task of learning a planning action model from {\em state-invariants} as a tuple $\Lambda=\tup{\mathcal{M},{\mathcal O},\Phi}$, where:
\begin{itemize}
\item $\mathcal{M}$ is the {\em initial empty model} that contains only the name, $name(\xi)$, and parameters, $pars(\xi)$, of each action model $\xi$ to be learned.
\item $\mathcal{O}=\tup{s_0^o,s_1^o \ldots, s_m^o}$ is a single learning example. This example can be reduced to its minimal expression $\mathcal{O}^*=\tup{s_0^o, s_m^o}$ meaning that it only contains two state observations, a full initial state $s_0^o$ and a partially observed state $s_m^o$. Note that the set of predicates $\Psi$ is deducible from $\mathcal{O}$ since $s_0^o$ is a fully observed state.
\item $\Phi$ is a set of {\em state-invariants} that define constraints about the set of possible states.
\end{itemize}

$M$ is the {\em space of possible action models} for the $A[\cdot]$ actions (i.e., the set of possible specifications of the $\rho$ and/or $\theta$ functions for each $a\in A[\cdot]$ action). We say that a given model $\mathcal{M}\in M$ is a {\em solution} to the $\Lambda=\tup{\mathcal{M},{\mathcal O},\Phi}$ learning task iff there exists a plan $\pi$ that solves $P=\tup{F,A[\cdot],I,G}$, when the semantics of each action $a\in A[\cdot]$ is given by $\mathcal{M}$, and such that any state traversed by a trajectory $\tau(\pi,P)$ is {\em consistent} with the input set of {\em state-invariants} $\Phi$.

Now, we show how to exploit our compact encoding of \strips\ action models to solve a $\Lambda=\tup{\mathcal{M},{\mathcal O},\Phi}$ learning task with an off-the-shelf classical planner.

\section{Learning action models from {\em state-invariants} with classical planning}
\label{sec:compilation}
Given a $\Lambda=\tup{\mathcal{M},{\mathcal O},\Phi}$ learning task we build and solve a classical planning problem $P_{\Lambda}=\tup{F_{\Lambda},A_{\Lambda},I,G_{\Lambda}}$ such that:
\begin{itemize}
\item $F_{\Lambda}$ extends $F$ with a fluent {\small$mode_{inval}$}, to indicate whether an action model is {\em inconsistent} with the input {\em state-invariants} $\Phi$, a fluent {\small$mode_{insert}$}, to indicate whether action models are being programmed, and the fluents for the propositional encoding of the corresponding space of STRIPS action models. As explained, this is a set of fluents of the type $\{pre\_e\_\xi, eff\_e\_\xi\}_{\forall e\in{\mathcal I}_{\Psi,\xi}}$.

@@

\subsection{Pruning inconsistent action models with {\em domain mutex}}
Our approach to implement this pruning is extending the conditional effects of the $\mathsf{insertPre_{p,\xi}}$ and $\mathsf{insertPre_{p,\xi}}$ actions (i.e., the actions that determine a solution model $\mathcal{M}$) with extra conditional effects indicating that the programmed model is {\em invalid} (i.e., inconsistent with a {\em domain mutex} in $\Phi$). Note that this {\em consistency} checking is more effective than the one implemented at the $\mathsf{apply_{\xi,\omega}}$ actions since $\mathsf{insertPre_{p,\xi}}$ and $\mathsf{insertPre_{p,\xi}}$ actions appear at an earlier stage of the planning process.

>>>>

\begin{abstract}
This paper addresses the learning of \strips\ action models from {\em state-invariants} (i.e logic formulae that specify constraints about the possible states of a given domain). The benefit of exploiting {\em state-invariants} is two-fold, they constrain the space of possible action models and they can complete learning examples that are only partially observed. Our approach for learning \strips\ action from {\em state-invariants} is a {\em classical planning} compilation that is flexible to different sources of input knowledge including partial observations of plan executions and {\em state-invariants}. The experimental results demonstrate that, even at unfavorable scenarios where input observations are minimal (a single learning example that comprises just a full initial state and a partially observed state), {\em state-invariant} are helpful to learn good quality \strips\ action models.
\end{abstract}

\section{Introduction}
\label{sec:introduction}
The specification of action models is a complex process that limits, too often, the application of {\em model-based planning} to real-world tasks~\cite{kambhampati:modellite:AAAI2007}. The {\em machine learning} of action models relieves this {\em knowledge acquisition bottleneck} of {\em model-based planning} and nowadays, there exists a wide range of effective approaches for learning action models~\cite{arora:amodels:ker2018}. Many of the most successful approaches for learning planning action models are however purely {\em inductive}~\cite{yang2007learning,pasula2007learning,mourao2010learning,zhuo2013action}, linking learning performance exclusively to the {\em amount} and {\em quality} of the input learning examples (which typically are observation of plan executions).

This paper addresses the learning of action models exploiting a different source of knowledge, {\em deductive} knowledge. Our approach leverages {\em state-invariants} (i.e. logic formulae that specify constraints about the possible states of a given domain) to cushion the negative impact of insufficient learning examples. Given an action model, state-of-the-art planners infer {\em state-invariants} from that model to reduce the search space and make the planning process more efficient~\cite{helmert2009concise}. In this paper we follow the opposite direction and leverage {\em state-invariants} to learn the planning action model. The benefit of learning action models from {\em state-invariants} is two-fold, {\em state-invariants} constrain the space of possible action models and can complete learning examples that are only partially observed.

Our approach for learning \strips\ action models from {\em state-invariants} is to compile the learning task into a classical planning task. Our compilation is built on top of the classical planning compilation for the learning of \strips\ action models~\cite{aineto2018learning} and it is flexible to different kinds of input knowledge including both partial observations of plan executions and {\em state-invariants}. The compilation outputs an \strips\ action model that is {\em consistent} with all the given input knowledge. The experimental results demonstrate that, even at unfavorable scenarios where input observations are minimal (a single learning example that comprises just a full initial state and a partially observed state), {\em state-invariant} are helpful to learn good quality \strips\ action models.

\section{Background}
\label{sec:background}
This section formalizes the {\em planning model} we follow in this work and introduces the classical planning compilation for the learning of \strips\ action models~\cite{aineto2018learning}. Finally, the section formalizes {\em state-invariants}.

@@

\subsection{Learning action models with classical planning}
The {\em classical planning compilation} for the learning \strips\ action models~\cite{aineto2018learning} receives as input an empty model $\mathcal{M}$ (which contains just the action headers), and an observation of a set of observations of plan executions. The compilation outputs a model $\mathcal{M'}$ that specifies the preconditions and effects of each action schema included in $\mathcal{M}$ such that the validation of the observations following $\mathcal{M'}$ is successful; i.e., it holds $\rho(s_{i-1}^o,a_i)$ for every observed action and $s_i^o=\theta(s_{i-1}^o,a_i)$ for every observed state.

A solution plan to the classical planning problem that results from the compialtion is a sequence of: (a) \emph{insert actions}, that insert preconditions and effects on a schemata of $\mathcal{M}$ to build $\mathcal{M'}$ and (b) \emph{apply actions} that validate the application of the $\mathcal{M'}$ model in the input observations. Figure~\ref{fig:plan-lplan} shows a solution to a classical planning problem resulting from the~\citeauthor{aineto2018learning}~\citeyear{aineto2018learning} compilation corresponding to the {\em blocksworld}~\cite{slaney2001blocks}. In the initial state of that problem the robot hand is empty and three blocks (namely {\small\tt blockA}, {\small\tt blockB} and {\small\tt blockC}) are clear and on top of the table. The problem goal is having the three-block tower {\tt blockA} on top of {\tt blockB} and {\tt blockB} on top of {\tt blockC}. The plan shows the {\em insert} actions for the {\tt\small stack} scheme (steps $00-01$ insert the preconditions, steps $05-10$ insert the effects), the plan steps $02-04$ that  insert the preconditions of the {\tt\small pickup} scheme and steps $10-13$ that insert the effects of this scheme. Finally, steps $14-17$ is a plan postfix with actions that apply the programmed model to achieve the goals starting from the given initial state.

\begin{figure}[hbt!]
	{\tiny\tt
\begin{tabular}{ll}
		{\bf 00}:(insert\_pre\_stack\_holding\_v1) & {\bf 10}:(insert\_eff\_pickup\_clear\_v1) \\
		01:(insert\_pre\_stack\_clear\_v2) & 11:(insert\_eff\_pickup\_ontable\_v1)\\
                {\bf 02}:(insert\_pre\_pickup\_handempty) & 12:(insert\_eff\_pickup\_handempty)\\
                03:(insert\_pre\_pickup\_clear\_v1) & 13:(insert\_eff\_pickup\_holding\_v1)\\
                04:(insert\_pre\_pickup\_ontable\_v1) & {\bf 14}:(apply\_pickup blockB)\\
                {\bf 05}:(insert\_eff\_stack\_clear\_v1) & 15:(apply\_stack blockB blockC)\\
                06:(insert\_eff\_stack\_clear\_v2) & 16:(apply\_pickup blockA)\\
                07:(insert\_eff\_stack\_handempty) & 17:(apply\_stack blockA blockB) \\
                08:(insert\_eff\_stack\_holding\_v1) &  {\bf 18}:(validate\_1)\\
                09:(insert\_eff\_stack\_on\_v1\_v2) &             		 
\end{tabular}
}
	\caption{\small Example of a solution to a problem output by the classical planning compilation for the learning \strips\ action models.}
	\label{fig:plan-lplan}
\end{figure}

@@

{\em State invariants} is a kind of state-constraints useful for computing more compact state representations~\cite{helmert2009concise} or making {\em satisfiability planning} and {\em backward search} more efficient~\cite{rintanen2014madagascar,alcazar2015reminder}. Given a classical planning problem $P=\tup{F,A,I,G}$, a {\em state invariant} is a formula $\phi$ that holds at the initial state of a given classical planning problem, $I\models \phi$, and at every state $s$, built from $F$, that is reachable from $I$ by applying actions in $A$. For instance, Figure~\ref{fig:strongest-invariant} shows five clauses that define {\em state invariants} for the {\em blocksworld} planning domain.

@@

@@

\subsection{The sampling space}
We define a {\em learning example} as a sequence $\mathcal{O}=\tup{s_0^o,s_1^o \ldots, s_m^o}$ of partial states, except for the initial state $s_0^o$, which is fully observed. Intermediate states can be missing meaning that are {\em unobserved}, so transiting between two consecutive observed states in $\mathcal{O}$ may require the execution of more than a single action ($\theta(s_i^o,\tup{a_1,\ldots,a_k})=s_{i+1}^o$, where ${\small k\geq 1}$ is unknown but finite. A partially observed state $s_i^o$ is one in which $|s_i^o| < |F|$; i.e., a state in which at least a fluent of $F$ is not observable. Note that this definition also comprises the case $|s_i^o| = 0$, when the state is fully unobservable.

@@

\subsection{The learning task}
We define the task of learning a planning action model from {\em state-invariants} as a tuple $\Lambda=\tup{\mathcal{M},{\mathcal O},\Phi}$, where:
\begin{itemize}
\item $\mathcal{M}$ is the {\em initial empty model} that contains only the name, $name(\xi)$, and parameters, $pars(\xi)$, of each action model $\xi$ to be learned.
\item $\mathcal{O}=\tup{s_0^o,s_1^o \ldots, s_m^o}$ is a single learning example. This example can be reduced to its minimal expression $\mathcal{O}^*=\tup{s_0^o, s_m^o}$ comprissing at least two state observations, a full initial state $s_0^o$ and a partially observed state $s_m^o$. The set of predicates $\Psi$ that shape the setf of fluents $F$ is deducible from $\mathcal{O}$ since $s_0^o$ is a fully observed state.
\item $\Phi$ is a set of {\em state-invariants} that define constraints about the set of possible states.
\end{itemize}

Next section shows how to exploit our compact encoding of \strips\ action models to solve a $\Lambda=\tup{\mathcal{M},{\mathcal O},\Phi}$ learning task with an off-the-shelf classical planner.

\section{Learning action models from {\em state-invariants} with classical planning}
\label{sec:compilation}

Given a $\Lambda=\tup{\mathcal{M},{\mathcal O},\Phi}$ learning task we build and solve a classical planning problem $P_{\Lambda}=\tup{F_{\Lambda},A_{\Lambda},I,G_{\Lambda}}$ such that:
\begin{itemize}
\item $F_{\Lambda}$ extends $F$ with a fluent {\small$mode_{inval}$}, to indicate whether an action model is {\em inconsistent} with the input {\em state-invariants} $\Phi$, a fluent {\small$mode_{insert}$}, to indicate whether action models are being programmed, and the fluents for the propositional encoding of the corresponding space of STRIPS action models. As explained, this is a set of fluents of the type $\{pre\_e\_\xi, eff\_e\_\xi\}_{\forall e\in{\mathcal I}_{\Psi,\xi}}$.

@@

\subsection{Pruning inconsistent action models with {\em domain mutex}}
We define a {\em domain mutex} as a $\tup{p,q}$ predicates pair where both $p\in\Psi$ and $q\in\Psi$ are predicates that shape the set of fluents $F$ of a given planning problem and such that they satisfy the following formulae $p\rightarrow \neg q$ where the predicate variables are universally quantified. For instance, predicates $holding(x)$ and $clear(x)$ from the {\em blocksworld} are {\em domain mutex} while predicates $clear(x)$ and $ontable(x)$ are not ($\forall x\ clear(x)\leftrightarrow\neg ontable(x)$ does not hold for every possible {\em blocksworld} state). We pay special attention to {\em domain mutex} because they identify the {\em properties} of a given type of objects~\cite{fox:TIM:JAIR1998} and because they enable (1) effectively pruning of inconsistent \strips\ action models and (2) effective completion of partially observed states.

Our approach to implement this pruning is extending the conditional effects of the $\mathsf{insertPre_{p,\xi}}$ and $\mathsf{insertPre_{p,\xi}}$ actions (i.e., the actions that determine a solution model $\mathcal{M}$) with extra conditional effects indicating that the programmed model is {\em invalid} (i.e., inconsistent with a {\em domain mutex} in $\Phi$). Note that this {\em consistency} checking is more effective than the one implemented at the $\mathsf{apply_{\xi,\omega}}$ actions since $\mathsf{insertPre_{p,\xi}}$ and $\mathsf{insertPre_{p,\xi}}$ actions appear at an earlier stage of the planning process.
